Are we learning to build better models at the cost of better decisions?

### 2021-03-08 manuscript/manuscript.Rmd
:pen:

abstract: |
  Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response. It is taken for granted that the model which makes the most accurate forecast, accounting for uncertainty, will also be the best model to inform decision-making.  Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.

@@

keywords:
- transients,
- optimal control,
- adaptive management,
- stochasticity,
- uncertainty,
- ecological management

@@

Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other. Importantly, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.  The manager quickly concludes that Model 1 is entirely implausible, but also finds Model 2 to be remarkably accurate at predicting future values, and is entirely unaware that the model is leading to such poor outcomes. A more diligent manager may seek to improve Model 2 further by re-estimating parameters, but this in fact leads to a model with slightly lower $K$ and thus even higher level of over-fishing and under-performance than Model 2. The only way to avoid this trap is not to learn with each new observation, not to apply iterative updates over the models. A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes.  This raises several questions.  Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? How would we know? And what should we do about it?

It is tempting to argue that this situation would almost never arise in real world scenarios.  In this argument, Model 1 is a unicorn, a mythical model that somehow gives optimal performance while at the same time making predictions that are wildly wrong.  Could such unicorns really exist?  This is a difficult assertion to prove either way, but it is worth noting that neither of these features by themselves are especially rare. Examples of serious forecasts that widely miss future observations abound wherever forecasts are common, from elections to economics to environmental change [@].  Meanwhile, the premise that model need not perfectly capture reality to be useful is at the very heart of modeling.  Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].  Capturing only the essential aspects as simply as possible is the goal of any model building exercise.  Moreover, rigorous updating of models and parameters through approaches such as adaptive management or iterative forecasting are not yet widespread practices in ecological management [@Dietze2018].  If such unicorn models exist, that lack of adoption would make it easier for them to persist at least until now.

If these unicorns do exist in real world management, then what do we do about it? Our step may be to acknowledging that we do not yet have the answer.  There are at least three common ideas which are not the answer, which might succinctly be termed (1) don't care, (2) don't learn or (3) don't model.

The first response, don't care, suggests that iterative updating is always desirable.  Previous work has acknowledged that future observations may be relatively uninformative, failing to reduce uncertainty between models by much.  This is the motivation for active over passive adaptive management, [@Walters1978], and has also acknowledge that even significant improvements in predictive accuracy may have negligible improvement on the quality of the decision.  This is the motivation for quantifying the value of information, [@Katz1987].  However, the conclusion from such observations is that additional information is at worst harmless to the quality of the decision, while in this example it is actively detrimental.  The realized 'value of information' contained in future measurements is negative.

The

Why then do our best available approaches so readily reject the model?

The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  We have no methods of reducing model uncertainty which do not perform significantly worse on this management problem than an approach which does not attempt to learn.  Put another way, the value of information contained in future observations [@] is always negative in this example.

So why do our best approaches fail such a simple test case? It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds:   (3) Don't we already know this?

(1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved. I address these in turn.

@@

This issue is not unique to sequential decision problems or optimal control solutions.  For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.  This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.  Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem. Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2. Does this mean MSE is not suseptible to this issue?  No, it does not.  It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.  Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954]. If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.

Is this result an artifact of the simple models considered here?  A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980].  Such complexity would not be sufficient to resolve the paradox of model uncertainty illustrated here. Simplicity facilitates intuition: Model 1 performs best because it correctly approximates the stock size which achieves maximum growth rate under the true model.  In contrast, each of those widely used recruitment relationships are symmetric, and their best-fitting parameter estimates will always have a maximum growth rate at a lower stock size. It is easy to point out that even this wider array of parameter and model values is too simple, since fisheries models in practice are much more complicated still (indeed, too complicated for the formal integration over model uncertainty considered here).  It is indeed true that either the forecasting or adaptive management example would work as expected if the set of candidate models included the true model, or at least a functional form that could better approximate the peak of the true model while also fitting better than model 2.  But we never have the true model in real systems.   If the candidate models are only caricatures, so too is the "true" model.  The candidate models considered here are no doubt much closer approximations to model 3 (all being single species, unstructured in age, stage, time or space, etc) than any model used in practice is to reality. Morever, a more accurate model is not required to reach optimal outcomes -- Model 1 already provides this by correctly approximating the key feature in this problem, even while it is wrong about much else.  How do proceed when our best methods fail in such a simple case?

>>>>

abstract: |
  Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response. Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.

@@

keywords:
- forecasting,
- adaptive management,
- stochasticity,
- uncertainty,
- optimal control

@@

Given this simple decision problem in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other. Moreover, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.  In both approaches, the manager quickly concludes that Model 1 is entirely implausible, while finding that Model 2 is remarkably accurate at predicting future values. This problem is not addressed by re-estimating parameters (equivalently, considering a larger suite of models with all possible $K$ and $r$ values): in fact leads to a model with slightly lower $K$ and even higher level of over-fishing and under-performance than Model 2 (Appendix B, Fig S1).

The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.  Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.  While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.  This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.

This phenomenon is not unique to sequential decision problems or optimal control solutions.  For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.  This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.  Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem. Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2. Does this mean MSE is not susceptible to this issue?  No, it does not.  It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.  Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954]. If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.

Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, which also involves a better-fitting model leading to worse outcomes [].  In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.  The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy. But in the example here, both Model 1 and Model 2 have the same structural complexity.  The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit.

Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial, but has failed to recognize the possibility that it can potentially be detrimental.  For example, active adaptive management is premised on the observation that future observations may be too uninformative to distinguish between alternatives [@Walters1978].  Others have also noted that even big improvements in predictive accuracy could have negligible improvement on the decision outcomes, which motivates quantifying the Value of Information (VOI) [@Katz1987]. But in both cases, the worse outcome is wasted effort.  In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.  The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates. A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes than alternative models.  This raises several questions.  Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? If so, How would we know, and what should we do about it? I address these in turn.

<!-- **Is it just a unicorn?** -->
Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?  It is tempting to argue that Model 1 is a kind of unicorn, a mythical creature existing only in theory. But that is a difficult assertion to prove.  Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [@].  If there is no shortage of bad forecasts, then could any of them be useful?  The premise that model need not perfectly capture reality to be useful is at the very heart of modeling.  Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].  Capturing only the essential aspects as simply as possible is the goal of any model building exercise.  Thus it should be no surprise that a model can drive good decisions while making poor predictions.  It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management or iterative forecasting [@Dietze2018].  This makes it more likely that such models could persist in practice until now. While iterations that revise this models will no doubt improve decision outcomes in many case, it is worth bearing in mind from the example here that such a connection is not guaranteed.

<!-- So what? -->
If these unicorns do exist in real world management, then what do we do about them? I believe this is an open question, but that our first step must be to recognize it as such.  I contrast that position with two common but diametrically opposed objections: (1) that the issue is easily resolved by including more or better models, or (2) this issue is easily resolved by relying less on models.  These are both important points.

The union of two simple models considered here (Models 1 & 2) certainly fails to fully capture the inherent uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [].  Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations.   However, it must also be noted that we could have included a much more realistic approach of uncertainty over many possible values of $r$, $K$ (see Appendix B, Fig S1), or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], without altering the results at all (these models also all have symmetric growth rates with a peak growth rate at half the steady-state population size).  Models will always be simpler than reality.  Surely it would be preferable if our methods could prefer the nearly optimally-performing Model 1 which we already have than require the issue be solved only by adding more and more models?

If we cannot always resolve the issue by adding more models, is this just another example of the risks of an over-reliance on models at all?

@@
